Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-03-23 NovelLBM.tex
units

\usepackage{hyperref} %link support
\hypersetup{colorlinks,breaklinks,
            allcolors=blue}

@@

	\begin{center}
		$^1$ Biofluid Simulation and Modeling – Theoretische Physik VI, Universität Bayreuth, Bayreuth, Germany\\
		$^2$ Department of Functional Materials in Medicine and Dentistry Institute of Functional Materials and Biofabrication (IFB) and KeyLab Polymers for Medicine of the Bavarian Polymer Institute (BPI) Universität Würzburg, Würzburg, Germany
	\end{center}
	\vspace{\fill}
	\begin{center}
		%TODO abstact
	\end{center}
	%TODO keywords

@@

The PTT model can be written in the form of a differential equation for the polymer stress $\ma{\tau}$ \cite{ferras2019}:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
We note that \cite{ferras2019} have also studied extended forms of the PTT model with a Mittag-Leffler instead of an exponential function, whereas here we restrict ourselves to the exponential version.
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at which shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \vec{u}\cdot\nabla\ma{A}.
	\end{equation}
\SG{ Kannst du der Vollständigkeit halber nochmal die Definition von $ \nabla \ma{ A }$ hinschreiben? }	
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function.
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.8e4} \\
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc} and corresponding viscosity ratios from eq.~(\ref{eq:R_eta}).}
		\label{tab:PTTparameters}
	\end{table}

@@

\begin{table}
\begin{center}
			\begin{tabular}{l|r|r|l}
				Author & max(Wi) & max($R_\eta$) & $R_\eta$ estimation \\
				\hline
				Malaspinas\cite{malaspinas2010} & 10 & 9 & In 4.3: $R_\nu = \num{0.1}$ \& equation (38) \\
				\rowcolor{secondaryBackground}
				Gupta\cite{gupta2015} & 100 & 0.7 & In IV.B: $\frac{\eta_\text{p}}{\eta_\text{A} + \eta_\text{p}} = \num{0.4}$ \\
%				Dzanic\cite{dzanic2022} & 10 & 0.5 & In 4.1: $\beta = \frac{\nu_\text{p}}{\nu_\text{s}} = \num{0.5}$ \\
				Ma\cite{ma2020} & 100 & 9 & In 4.1: $\beta = \frac{\mu_\text{s}}{\mu_\text{s} + \mu_\text{p}} = \num{0.1}$ \\
				\rowcolor{secondaryBackground}
				Onishi\cite{onishi2006} & ? & 1 & In 3.1: $\beta = \frac{\mu_\text{p}}{\mu_\text{s} + \mu_\text{p}}$ \& Table 1: $\beta=\num{0.5}$ \\
				Wang\cite{wang2020} & \glqq up to O(1)\grqq & 5 & In II: $\eta_\text{p} = c\eta_\text{s}$ \& FIG. 9: $c = \num{5}$\\
				\rowcolor{secondaryBackground}
%				Dzanic\cite{dzanic2022_2} & 10 & 9 & In 5.1: $\beta = \frac{\nu_\text{s}}{\nu_\text{s} + \nu_\text{p}}$ \& In 5.1.2: $\beta = \num{0.1}$ \\
				Su\cite{su2013_2} & 10 & 1  & In II: $\beta = \frac{\eta_\text{s}}{\eta_\text{s} + \eta_\text{p}}$ \& In IV.A: $\beta = \num{0.5}$ \\
				Gupta\cite{gupta2015_2} & ? & 0.7 & In 2: $\frac{\eta_\text{p}}{\eta_\text{d}} = \num{0.4}$ \& $\eta_\text{d} = \eta_\text{A} + \eta_\text{p}$ \\
				\rowcolor{secondaryBackground}
				Kuron\cite{kuron2021} & 1 & 9 & In 4.1: $\beta = \num{0.9}$ \& equations (14) \& (15) \\
				Gupta\cite{gupta2016} & 80 & 0.7 & In 3: $\frac{\eta_\text{p}}{\eta_\text{c,d} + \eta_\text{p}} = \num{0.4}$ \\
				\rowcolor{secondaryBackground}
				Onishi\cite{onishi2005} & 1000 & 1 & In 3.2: $\frac{\eta_\text{p}}{\eta_\text{s}} = \num{1}$ \\
			\end{tabular}
		\end{center}

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$ here and check different values further below. 
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched. For this we calculate $\tau_{xx}$ and $\tau_{xy}$ using equations~(\ref{eq:tau_xx}) and (\ref{eq:PTTDerivation2}) using the $\dot{\gamma}$ calculated from equation~(\ref{eq:semiAnalyticGD_3D}). The polymer stress is then initialized using these values. 
%\SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\RK{done}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
The time evolution of the center-line velocity is shown in figure~\ref{fig:2DTimeEvolution} of the appendix.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
Using the simulated velocities $\vec{u_i}$ and the semi-analytical results $\vec{u}$. This gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved L1 error defined as follows is shown in figure~\ref{fig:2DvelErr} in the appendix.
\begin{equation}
	e_\text{L1} = \frac{\vec{u_i} - \vec{u}\left(y_i\right)}{\vec{u}\left(y_i\right)}
\end{equation}
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.
Similarly, the exact value of $\alpha_\text{s}$ does not matter as can be seen in figure~\ref{fig:ShuffleInPoiseuille}.

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of $40 \times 40 \mu$m. %\qtyproduct{40x40}{\micro\meter}.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal}. \SG{ units. Is this Pa/m or is this the pressure difference between inlet and outlet? }
In figure~\ref{fig:RT-DC}, we show the velocity profile, the von Mises stress and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find a very low and a very high viscosity respectively. These can be outside the range of viscosities, that can be obtained through rheological measurements with the fluid. This effect is even more pronounced along the walls directly after the outlet. Here the viscosities obtained in this manner become negative. Due to the strain rate approaching $0$ at several locations within the simulation volume, this calculation often diverges. The clamped result can be seen in figure~\ref{fig:Nozzle/RT-DC_eta}. These discrepancies, especially the ones at the walls, show the breakdown of the viscosity interpretation in viscoelastic cases and signify the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by purely viscous models and should be considered despite the high computational cost.

@@

The solutions were prepared by dissolving alginate (DuPont VIVAPHARM Alginate PH176) in ultrapure water at a ratio of 4 \% w/v. 
To ensure homogeneous dissolution, the material was stored at 37$^\circ$C for 24 h and regularly mixed to ensure complete dissolution. 
The measurements were conducted with a 25 mm plate-plate geometry at a 500 $\mu$m gap on a rheometer (Anton paar MCR 702). 
A solvent trap was used to limit evaporation. 
All measurements were done at a set temperature of 25$^\circ$C as shear rate sweep. 
Data is agglomerated from 21 independent measurements.
The first normal stress difference $N_1$ is calculated from the measured normal force using the following equation \cite{kulicke1977}:
	\begin{equation}
		N_1 = \frac{2F}{\pi R^2} + \frac{3}{20}\rho\omega^2R^2
	\end{equation}
	where $F$ is the measured normal force, $R$ is the radius of the rheometer, $\rho$ is the density of the fluid and $\omega$ is the angular velocity of the rheometer. The second term is usually omitted and as we found it to matter little, and as the quality of the data we have available is too limited to warrant such detailed corrections, we also choose to omit it. A cone-rheometer would be preferred here, but as $N_2 = 0$ for PTT with $\xi = 0$, the difference is negligible.

@@

As additional quantitative validation, we simulate a pure shear flow driven by imposing a fixed velocity at the top and bottom of the computational box and using periodic boundary conditions in the other two directions.
The box dimension is 2x43x2.
We use PTT with the methyl cellulose parameters and vary the shear rate to cover the full range starting from the zero-shear all the way to the infinite-shear plateau displayed in figure~\ref{fig:fit_mc} of the main text.
The shuffling is fixed at $\alpha_\text{s} = 1$ and the viscosity as function of shear rate is computed by dividing the total fluid stress at the center of the box by the shear rate.
Figure~\ref{fig:wiParameterStudy}~(a) shows that our simulations are in very good agreement with the semi-analytical solution with a maximum error of \num{2e-11}.
Importantly, we show in figure~\ref{fig:wiParameterStudy}~(b) that this very good agreement is independent of the shuffling parameter $\alpha_\text{s}$.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/wiParameterStudy.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/suffleParameterStudy.pdf}
		\caption{(a) The local viscosity determined by a shear flow simulation is in very good agreement with the semi-analytical theory.
		(b) The error is negligibly small and independent of the shuffling parameter $\alpha_\text{s}$. \SG{ at which $ \dot{ \gamma  }$? }}
		\label{fig:wiParameterStudy}
	\end{figure}

@@

		\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/ShuffleInPoiseuille.pdf}
		\caption{In Poiseuille flow, the result is near identical for different $\alpha_\text{s}$. The error reveals minor differences. Aside from the wall nodes, the largest error appears for small shuffle fractions. For large ones, the error appears to become independent of $\alpha_\text{s}$. This indicates, that the observed difference might be due to the Mach number changing for different $\alpha_\text{s}$ and not due to the shuffling itself. This is due lower $\alpha_\text{s}$ leading to lower LBM viscosities and such to higher Mach numbers. \SG{ is this planar or cylindrical? }}
		\label{fig:ShuffleInPoiseuille}
	\end{figure}

>>>>

\usepackage{hyperref} %link support
\hypersetup{colorlinks,breaklinks,allcolors=blue}

@@

	\begin{center}
		$^1$ Biofluid Simulation and Modeling – Theoretische Physik VI, Universität Bayreuth, Bayreuth, Germany\\
		$^2$ Department of Functional Materials in Medicine and Dentistry Institute of Functional Materials and Biofabrication (IFB) and KeyLab Polymers for Medicine of the Bavarian Polymer Institute (BPI) Universität Würzburg, Würzburg, Germany
	\end{center}
	\vspace{\fill}
	%TODO keywords

@@

The PTT model can be written in the form of a differential equation for the polymer stress $\ma{\tau}$ \cite{ferras2019}:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
We note that \cite{ferras2019} have also studied extended forms of the PTT model with a Mittag-Leffler instead of an exponential function, whereas here we restrict ourselves to the exponential version.
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at which shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \left(\vec{u}\cdot\nabla\right)\ma{A}.
	\end{equation}
\SG{ Kannst du der Vollständigkeit halber nochmal die Definition von $ \nabla \ma{ A }$ hinschreiben? }\RK{Nein}
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function.
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.82e4} \\
			\rowcolor{secondaryBackground}
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			\rowcolor{secondaryBackground}
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc} along with corresponding viscosity ratios from eq.~(\ref{eq:R_eta}).}
		\label{tab:PTTparameters}
	\end{table}

@@

\begin{table}
\begin{center}
			\begin{tabular}{l|r|r|l}
				Author & max(Wi) & max($R_\eta$) & $R_\eta$ estimation \\
				\hline
				Malaspinas\cite{malaspinas2010} & 10 & 9 & In 4.3: $R_\nu = \num{0.1}$ \& equation (38) \\
				\rowcolor{secondaryBackground}
				Gupta\cite{gupta2015} & 100 & 0.7 & In IV.B: $\frac{\eta_\text{p}}{\eta_\text{A} + \eta_\text{p}} = \num{0.4}$ \\
%				Dzanic\cite{dzanic2022} & 10 & 0.5 & In 4.1: $\beta = \frac{\nu_\text{p}}{\nu_\text{s}} = \num{0.5}$ \\
				Ma\cite{ma2020} & 100 & 9 & In 4.1: $\beta = \frac{\mu_\text{s}}{\mu_\text{s} + \mu_\text{p}} = \num{0.1}$ \\
				\rowcolor{secondaryBackground}
				Onishi\cite{onishi2006} & ? & 1 & In 3.1: $\beta = \frac{\mu_\text{p}}{\mu_\text{s} + \mu_\text{p}}$ \& Table 1: $\beta=\num{0.5}$ \\
				Wang\cite{wang2020} & "up to O(1)" & 5 & In II: $\eta_\text{p} = c\eta_\text{s}$ \& FIG. 9: $c = \num{5}$\\
				\rowcolor{secondaryBackground}
%				Dzanic\cite{dzanic2022_2} & 10 & 9 & In 5.1: $\beta = \frac{\nu_\text{s}}{\nu_\text{s} + \nu_\text{p}}$ \& In 5.1.2: $\beta = \num{0.1}$ \\
				Su\cite{su2013_2} & 10 & 1  & In II: $\beta = \frac{\eta_\text{s}}{\eta_\text{s} + \eta_\text{p}}$ \& In IV.A: $\beta = \num{0.5}$ \\
				Gupta\cite{gupta2015_2} & ? & 0.7 & In 2: $\frac{\eta_\text{p}}{\eta_\text{d}} = \num{0.4}$ \& $\eta_\text{d} = \eta_\text{A} + \eta_\text{p}$ \\
				\rowcolor{secondaryBackground}
				Kuron\cite{kuron2021} & 1 & 9 & In 4.1: $\beta = \num{0.9}$ \& equations (14) \& (15) \\
				Gupta\cite{gupta2016} & 80 & 0.7 & In 3: $\frac{\eta_\text{p}}{\eta_\text{c,d} + \eta_\text{p}} = \num{0.4}$ \\
				\rowcolor{secondaryBackground}
				Onishi\cite{onishi2005} & 1000 & 1 & In 3.2: $\frac{\eta_\text{p}}{\eta_\text{s}} = \num{1}$ \\
			\end{tabular}
		\end{center}

@@

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$ here and check different values further below.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched. For this we calculate $\tau_{xx}$ and $\tau_{xy}$ using equations~(\ref{eq:tau_xx}) and (\ref{eq:PTTDerivation2}) using the $\dot{\gamma}$ calculated from equation~(\ref{eq:semiAnalyticGD_3D}). The polymer stress is then initialized using these values.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
The time evolution of the center-line velocity is shown in figure~\ref{fig:2DTimeEvolution} of the appendix.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
Using the simulated velocities $\vec{u_i}$ and the semi-analytical results $\vec{u}$. This gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved L1 error defined as follows is shown in figure~\ref{fig:2DvelErr} in the appendix.
\begin{equation}
	e_\text{L1} = \frac{\vec{u_i} - \vec{u}\left(y_i\right)}{\vec{u}\left(y_i\right)}
\end{equation}
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.
Similarly, the exact value of $\alpha_\text{s}$ does not matter as can be seen in figure~\ref{fig:ShuffleInPoiseuille}.

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of \qtyproduct{40x40}{\micro\meter}.
%$\SI{40}{\micro\meter} \times \SI{40}{\micro\meter}$.%Correctly typeset version, that compiles for Stephan.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal\per\meter}. \SG{ units. Is this Pa/m or is this the pressure difference between inlet and outlet? }\RK{done}
In figure~\ref{fig:RT-DC}, we show the velocity profile, the von Mises stress and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find a very low and a very high viscosity respectively. These can be outside the range of viscosities, that can be obtained through rheological measurements with the fluid. This effect is even more pronounced along the walls directly after the outlet. Here the viscosities obtained in this manner become negative. Due to the strain rate approaching $0$ at several locations within the simulation volume, this calculation often diverges. The clamped result can be seen in figure~\ref{fig:Nozzle/RT-DC_eta}. These discrepancies, especially the ones at the walls, show the breakdown of the viscosity interpretation in viscoelastic cases and signify the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by purely viscous models and should be considered despite the high computational cost.

@@

The solutions were prepared by dissolving alginate (DuPont VIVAPHARM Alginate PH176) in ultrapure water at a ratio of \SI{40}{\milli\gram\per\milli\liter}, which here is refereed to as \SI{4}{\percent}.
To ensure homogeneous dissolution, the material was stored at \SI{37}{\celsius} for \SI{24}{\hour} and regularly mixed to ensure complete dissolution.
The measurements were conducted with a \SI{25}{\milli\meter} plate-plate geometry at a \SI{500}{\micro\meter} gap on a rheometer (Anton paar MCR 702).
A solvent trap was used to limit evaporation.
All measurements were done at a set temperature of \SI{25}{\celsius} as shear rate sweep.
Data is agglomerated from 21 independent measurements.
The first normal stress difference $N_1$ is calculated from the measured normal force using the following equation \cite{kulicke1977}:
	\begin{equation}
		N_1 = \frac{2F}{\pi R^2} + \frac{3}{20}\rho\omega^2R^2
	\end{equation}
	where $F$ is the measured normal force, $R$ is the radius of the rheometer, $\rho$ is the density of the fluid and $\omega$ is the angular velocity of the rheometer. The second term is usually omitted and as we found it to matter little, and as the quality of the data we have available is too limited to warrant such detailed corrections, we also choose to omit it. A cone-rheometer would be preferred here, but as $N_2 = 0$ for PTT with $\xi = 0$, the difference is negligible.

@@

As additional quantitative validation, we simulate a pure shear flow driven by imposing a fixed velocity at the top and bottom of the computational box and using periodic boundary conditions in the other two directions.
The box dimension is 2x43x2.
We use PTT with the methyl cellulose parameters and vary the shear rate to cover the full range starting from the zero-shear all the way to the infinite-shear plateau displayed in figure~\ref{fig:fit_mc} of the main text.
The shuffling is fixed at $\alpha_\text{s} = 1$ and the viscosity as function of shear rate is computed by dividing the total fluid stress at the center of the box by the shear rate.
Figure~\ref{fig:wiParameterStudy}~(a) shows that our simulations are in very good agreement with the semi-analytical solution with a maximum error of \num{2e-11}.
Importantly, we show in figure~\ref{fig:wiParameterStudy}~(b) that this very good agreement is independent of the shuffling parameter $\alpha_\text{s}$.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/wiParameterStudy.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/suffleParameterStudy.pdf}
		\caption{(a) The local viscosity determined by a shear flow simulation is in very good agreement with the semi-analytical theory.
		(b) The error is negligibly small and independent of the shuffling parameter $\alpha_\text{s}$. \SG{ at which $ \dot{ \gamma  }$? }\RK{\SI{1}{\per\second}}}
		\label{fig:wiParameterStudy}
	\end{figure}

@@

		\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/ShuffleInPoiseuille.pdf}
		\caption{In planar Poiseuille flow, the result is near identical for different $\alpha_\text{s}$. The error reveals minor differences. Aside from the wall nodes, the largest error appears for small shuffle fractions. For large ones, the error appears to become independent of $\alpha_\text{s}$. This indicates, that the observed difference might be due to the Mach number changing for different $\alpha_\text{s}$ and not due to the shuffling itself. This is due lower $\alpha_\text{s}$ leading to lower LBM viscosities and such to higher Mach numbers. \SG{ is this planar or cylindrical? }\RK{done}}
		\label{fig:ShuffleInPoiseuille}
	\end{figure}
